Supporting Information:\\Black-swan events in animal populations

### 2016-05-27 text/ms.tex
Additions to main text

@@

There are a number of possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a t distribution with heavy
tails \citep{allen2001}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black swans, or sharp changes in human
pressures such hunting, fishing, or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.

@@

We think modelers should routinely consider representing population dynamic
process noise with a (possibly downward-skewed)
heavy-tailed distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to timeseries
with suspected mass mortality events
\citep[e.g.~the t distribution compared to a mixture
of two normal distributions;][]{ward2007}.
In the meantime, we think the t distribution
represents a simple solution.
It can be easily substituted for the normal distribution
and if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
then the normal distribution.
Barring a better estimate, we suggest $\nu = 4$ or $5$ and possibly a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a).
Above $\nu  =  10$ the t distribution closely resembles
the normal distribution,
and $\nu  = 4$ or $5$ allows for an event
three standard deviations below the mean to happen
once every 50--65 years instead of once every $\sim$ 740 years.
Adopting heavy-tailed process noise follows the intuition
that our models are usually reasonable but occasionally very wrong.

>>>>

We picked the t distribution mainly for pragmatic reasons:
it requires the estimation of only one additional
parameter over the normal distribution and reverts to the normal distribution
if strong evidence of heavy-tailed process deviations are not present.
In addition, the t distribution has some mechanistic justification.
The t distribution can be formed by a mixture of normal distributions
with the same mean and different variances \citep{gelman2014}.
This could represent the effect of unmodeled spatial heterogeneity,
a nonstationary environmental process, or
varying responses to temperature among individuals,
for example.

@@

We systematically evaluated the presence of black-swan events in hundreds of
animal populations. We find strong evidence for their occurrence. Black-swan
population crashes are substantially more frequent than black-swan population
increases and are usually driven by external events such as weather and
disease. Therefore, ignoring the presence of black-swan events, and the
predominance of crashes over increases, may underestimate extinction risk.

There are many possible causes of black-swan events, including
unmodelled intrinsic properties of populations or extrinsic forces acting on
populations. For example, we could observe black-swan dynamics if we miss an
underlying mixture of processes, since a mixture of normal distributions with
different variances can generate a t distribution with heavy
tails \citep{allen2001}. In ecology, these processes could arise through an
aggregation of populations across space, population
diversity \citep{schindler2010}, or an intrinsic change in population variability
through time \citep{carpenter2006}. Extrinsic ecological forces could also cause
black-swan dynamics \citep{nunez2012} including extreme climate
events \citep{meehl2004, katz2005, ipcc2012}, predation from (or competition
with) other species experiencing black swans, or sharp changes in human
pressures such hunting, fishing, or habitat destruction. Alternatively, the
synchrony of multiple extrinsic forces could give rise to black-swan dynamics
through synergistic interactions \citep{kirby2009} or a rare alignment of
non-synergistic forces \citep{denny2009}.

@@

% The models we fit may have misspecified
% the dynamics of the populations,
% but this is precisely the point of our paper.
% Black swans represent extreme deviations from our expectations
% and our expectations are represented by the models we fit.
% However, rarely, if ever, are our models perfect representations
% of the underlying truth.
% We might misunderstand the breeding biology of a species,
% the weather can suddenly change,
% disease can suddenly spread,
% and species can unexpectedly interact.

% todo the strongest correlates of finding blocks one events was the timeseries length
% ...implications

Given our results, we think modelers should routinely consider representing population dynamic
process noise with a heavy-tailed and downward-skewed distribution,
especially when making forecasts to evaluate risk.
But what heavy-tailed distribution should one pick
and how heavy tailed should it be?
% nearly any reason for why a population modeler has incorrectly specified the exact and full process generating the abundance times could generate a heavy tailed process noise.
This is an open research question.
The exact shape of the tails has proven important
in the field of dispersal biology \citep{kot1996, clark1999}.
Answers for the field of population dynamics
might come from simulation analyses exploring
the implications of different tail shapes and
from comparing the relative fit of heavy-tailed
population models to timeseries
with suspected mass mortality events
\citep[e.g.~the t distribution compared to a mixture
of two normal distributions;][]{ward2007}.
In the meantime, we think the t (or the skew-t) distribution
represents a simple solution.
It can be easily substituted for the normal distribution
and if the degrees of freedom parameter, $\nu$,
is fixed
it requires estimating no more parameters
then the normal distribution.
Barring a better estimate, we suggest $\nu = 4$ or $5$ and possibly a skewness parameter ($\gamma$) of $\sim 0.7$ (Fig.~4a).
Above $\nu  =  10$ the t distribution closely resembles
the normal distribution,
and $\nu  = 4$ or $5$ allows for an event
three standard deviations below the mean to happen
once every 50--65 years instead of once every $\sim$ 740 years.
Adopting heavy-tailed process noise follows the intuition
that our models are usually reasonable but occasionally very wrong.
